Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
todo in sec 2
Comments

The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
neutral terms and normal forms with respect to $\beta$-reduction\scw{Why not say parallel reduction here? We can be specific and say that that the only reductions available for these terms are identity reductions and cite
\footnote{\dotv{normalform.v}{nf\_refl}} }. Instead of the
metavariables $e$ and $f$, we also
use the judgment forms $[[ne a]]$ and $[[nf a]]$ to indicate that there exists
$[[e]]$ or $[[f]]$ such that $[[a = e]]$ or $[[a = f]]$.

@@

Because we are working with open terms, we need a few additional syntactic
lemmas about reduction. First, a renaming $\xi$ is a generalization of
weakening when working with de Bruijn indices. It consistently renumbers the
variables that appears in terms. If a renamed term has been reduced, we can
always recover the result of the reduction without the renaming.
%
\scw{This lemma is only used to prove \texttt{wn\_antirenaming}, which is then
used more generally. Maybe we should replace it with that?}
\begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
\end{lemma}

@@

Furthermore, to prove this adequacy property, we need to strengthen it to also
give us more information about neutral terms as we proceed by induction. In
particular, we need to know that all terms that reduce to neutral forms are
contained within the interpretation.
\scw{Also, cite where this form of proof originally
appeared? This part is standard.}
Therefore, we formally define when a set is a \emph{reducibility candidate}
(shortened as $CR$) as follows. Our definition of $CR$ is inspired by
\citet{girard1989proofs}, but not identical since we only care about
weak normalization.
\begin{definition}[Reducibility Candidates (CR)\footnote{\dotv{semtypingopen.v}{CR}}]
  Let $[[S]]$ be a set of lambda terms. We say that $[[S]] \in CR$
  if and only if conditions $CR_1$ and $CR_2$ hold.
  \begin{itemize}
  \item $[[S]] \in CR_1 \iff\ [[forall a, (#  wne a implies a in S #)]]$
  \item $[[S]] \in CR_2 \iff\ [[forall a, (# a in S implies wn a #)]]$
  \end{itemize}
\end{definition}

@@

\scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
sentence elsewhere? I'd like to use the previous sentence as a conclusion for
this section.}
Finally, due to technical
details, our mechanized proof about the existence of $\beta\eta$ form does not
immediately induce a decision procedure for type conversion in
\lang{}. We discuss this issue and its workaround in
Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ l |  c  | c | c }
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    % Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  68 & = & = \\
    Renaming & 46 & =  & = \\
    Untyped reduction & 349 & = & = \\
    Neutral and normal forms & N/A & 269 & N/A \\
    Logical relation & 331 & 426 & N/A \\
    Semantic typing and soundness & 169 & 199 & N/A \\
    Syntactic soundness & N/A & N/A  & 660 \\
    \hline
    Total & 963 & 1357 & 1123 \\
  \end{tabular}
  \end{minipage}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

@@

However, % with a little more effort,
it is possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal forms and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data.}.

\section{Related Work}
\label{sec:relatedwork}

@@

Second, the definition of our logical relation does
contribute to a more concise proof.
In \rref{I-Red, I-Bool}, we choose parallel reduction, a full
reduction relation, to close over our semantic interpretation of types
and terms. Parallel reduction is non-deterministic, but it satisfies
useful structural properties such as congruence
(Lemma~\ref{lemma:parcong}) and the diamond property
(Lemma~\ref{lemma:pardiamond}). We pay the price of using a
non-deterministic reduction relation when we want to prove that our
logical relation is a partial function; because of \rref{I-Red}, we
can have $[[A => B0]]$ and $[[A => B1]]$, where $[[B0]]$ and $[[B1]]$
each have their separate interpretations that we have to prove to be
equal. Fortunately, this complexity is reconciled by the
diamond property, which is easy to derive syntactically.
\citet{decagda,nbeincoq} employ a deterministic weak
head reduction relation. The use of a deterministic reduction relation
makes the functionality of a logical relation trivial to prove, but
the downside is that weak head reduction (denoted by $\leadsto^{wh}$)
or leftmost-outermost reduction (denoted by $\leadsto^{lo}$) does
not satisfy even the weaker substitution property
(Lemma~\ref{lemma:parsubst}).
With this alternative formulation, we would have to prove how the
deterministic reduction relates to a full non-deterministic reduction
relation to prove the fundamental theorem. This would amount
to proving the factorization theorem of the deterministic reduction
relation with respect to full non-deterministic reduction. Instead,
using parallel reduction in \rref{I-Red} allows us to delay the
factorization property until we need to extract a normalization
algorithm from our fundamental theorem (Section~\ref{sec:conversionalgo}).

>>>>

The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
neutral terms and normal forms with respect to $\beta$-reduction\scw{Why not say parallel reduction here? We can be specific and say that that the only reductions available for these terms are identity reductions and cite
\footnote{\dotv{normalform.v}{nf\_refl}} }.\yl{I always think of
normal form as the specific definition that says the relation can't
step. We can refer to nfrefl but that requires some explanation about
the definition of normal form and the complication that it doesn't
hold in $\eta$ (maybe it's fine if we just don't mention it in the
$\eta$ case)} Instead of the
metavariables $e$ and $f$, we also
use the judgment forms $[[ne a]]$ and $[[nf a]]$ to indicate that there exists
$[[e]]$ or $[[f]]$ such that $[[a = e]]$ or $[[a = f]]$.

@@

Because we are working with open terms, we need a few additional syntactic
lemmas about reduction. First, a renaming $\xi$ is a generalization of
weakening when working with de Bruijn indices. It consistently renumbers the
variables that appears in terms. If a renamed term has been reduced, we can
always recover the result of the reduction without the renaming.
%
\scw{This lemma is only used to prove \texttt{wn\_antirenaming}, which is then
used more generally. Maybe we should replace it with
that?}\yl{Anti-renaming is mentioned once in Section 6 when we talk abou how confluence with $\eta$
depends on anti-renaming for Par. I don't think that discussion in
Section 6 is very important so I'm fine with replacing it  }
\begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
\end{lemma}

@@

Furthermore, to prove this adequacy property, we need to strengthen it to also
give us more information about neutral terms as we proceed by induction. In
particular, we need to know that all terms that reduce to neutral forms are
contained within the interpretation.
Therefore, we formally define when a set is a \emph{reducibility candidate}
(shortened as $CR$) as follows. Our definition of $CR$ is inspired by
\citet{girard1989proofs}, but not identical since we only care about
weak normalization.
\begin{definition}[Reducibility Candidates (CR)\footnote{\dotv{semtypingopen.v}{CR}}]
  Let $[[S]]$ be a set of lambda terms. We say that $[[S]] \in CR$
  if and only if conditions $CR_1$ and $CR_2$ hold.
  \begin{itemize}
  \item $[[S]] \in CR_1 \iff\ [[forall a, (#  wne a implies a in S #)]]$
  \item $[[S]] \in CR_2 \iff\ [[forall a, (# a in S implies wn a #)]]$
  \end{itemize}
\end{definition}

@@

% \scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
% sentence elsewhere? I'd like to use the previous sentence as a conclusion for
% this section.}
% Finally, due to technical
% details, our mechanized proof about the existence of $\beta\eta$ form does not
% immediately induce a decision procedure for type conversion in
% \lang{}. We discuss this issue and its workaround in
% Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ l |  c  | c | c }
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    % Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  68 & = & = \\
    Renaming & 46 & =  & = \\
    Untyped reduction & 349 & = & = \\
    Neutral and normal forms & N/A & 269 & N/A \\
    Logical relation & 331 & 426 & N/A \\
    Semantic typing and soundness & 169 & 199 & N/A \\
    Syntactic soundness & N/A & N/A  & 660 \\
    \hline
    Total & 963 & 1357 & 1123 \\
  \end{tabular}
  \end{minipage}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

@@

However, % with a little more effort,
it is possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal forms and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data.}.
\section{Related Work}
\label{sec:relatedwork}

@@

Second, the definition of our logical relation does
contribute to a more concise proof.
In \rref{I-Red, I-Bool}, we choose parallel reduction, a full
reduction relation, to close over our semantic interpretation of types
and terms. Parallel reduction is non-deterministic, but it satisfies
useful structural properties such as congruence
(Lemma~\ref{lemma:parcong}) and the diamond property
(Lemma~\ref{lemma:pardiamond}). We pay the price of using a
non-deterministic reduction relation when we want to prove that our
logical relation is a partial function; because of \rref{I-Red}, we
can have $[[A => B0]]$ and $[[A => B1]]$, where $[[B0]]$ and $[[B1]]$
each have their separate interpretations that we have to prove to be
equal. Fortunately, this complexity is reconciled by the
diamond property, which is easy to derive syntactically.
\citet{decagda,nbeincoq} employ a deterministic weak
head reduction relation. The use of a deterministic reduction relation
makes the functionality of a logical relation trivial to prove, but
the downside is that weak head reduction (denoted by $\leadsto^{wh}$)
or leftmost-outermost reduction (denoted by $\leadsto^{lo}$) does
not satisfy even the weaker substitution property
(Lemma~\ref{lemma:parsubst}).
With this alternative formulation, we would have to prove how the
deterministic reduction relates to a full non-deterministic reduction
relation to prove the fundamental theorem. This would amount
to proving the factorization theorem of the deterministic reduction
relation with respect to full non-deterministic reduction. Instead,
using parallel reduction in \rref{I-Red} allows us to delay the
factorization property until we need to justify the use leftmost-outermost
reduction as a deterministic reduction strategy for normalization.
